---
title: "Zelig-bprobit"
output:
  html_document:
    number_sections: yes
    toc: yes
bibliography: zelig5choice.bib
---

<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{Model: Zelig-bprobit}
-->

# Bivariate Logistic Regression for Two Dichotomous Dependent Variables

Use the bivariate probit regression model if you have two binaryrun
dependent variables $(Y_1, Y_2)$, and wish to model them jointly as a
function of some explanatory variables. Each pair of dependent variables
$(Y_{i1}, Y_{i2})$ has four potential outcomes, $(Y_{i1}=1,
Y_{i2}=1)$, $(Y_{i1}=1, Y_{i2}=0)$, $(Y_{i1}=0, Y_{i2}=1)$, and
$(Y_{i1}=0, Y_{i2}=0)$. The joint probability for each of these four
outcomes is modeled with three systematic components: the marginal
Pr$(Y_{i1} = 1)$ and Pr$(Y_{i2} = 1)$, and the correlation parameter
$\rho$ for the two marginal distributions. Each of these systematic
components may be modeled as functions of (possibly different) sets of
explanatory variables.

## Syntax

```{r eval=FALSE, echo=FALSE}
z.out <- zelig(list(mu1 = Y1 ~ X1 + X2, 
                          mu2 = Y2 ~ X1 + X3,
                          rho = ~ 1),
                     model = "bprobit", data = mydata)
x.out <- setx(z.out)
s.out <- sim(z.out, x = x.out)
```

## Input Values

In every bivariate probit specification, there are three equations which
correspond to each dependent variable ($Y_1$, $Y_2$), and the
correlation parameter $\rho$. Since the correlation parameter does not
correspond to one of the dependent variables, the model estimates $\rho$
as a constant by default. Hence, only two formulas (for $\mu_1$ and
$\mu_2$) are required. If the explanatory variables for $\mu_1$ and
$\mu_2$ are the same and effects are estimated separately for each
parameter, you may use the following short hand:

```{r eval=FALSE}
fml <- list(cbind(Y1, Y2) ~ X1 + X2)
```

which has the same meaning as:

```{r eval=FALSE}
fml <- list(mu1 = Y1 ~ X1 + X2, + mu2 = Y2 ~ X1 + X2, rho = ~ 1)
```

You may use the function `tag()` to constrain variables
across equations. The `tag()` function takes a variable and a
label for the effect parameter. Below, the constrained effect of
`x3` in both equations is called the `age`
parameter:

```{r eval=FALSE}
fml <- list(mu1 = y1 ~ x1 + tag(x3, "age"), + mu2 = y2 ~ x2 + tag(x3, "age"))
```

You may also constrain different variables across different equations to
have the same effect.

## Examples

1.  Basic Example

    Load the data and estimate the model:
    
```{r echo = FALSE, warning = FALSE, message = FALSE}
suppressPackageStartupMessages(library(Zelig5Choice, quietly = TRUE))
data(sanction)
```

```{r}
data(sanction)
```

```{r}
z.out1 <- zelig(cbind(import, export) ~ coop + cost + target,
                model = "bprobit", data = sanction)
```

    By default, `zelig()` estimates two effect parameters for
    each explanatory variable in addition to the correlation
    coefficient; this formulation is parametrically independent
    (estimating unconstrained effects for each explanatory variable),
    but stochastically dependent because the models share a correlation
    parameter. Generate baseline values for the explanatory variables
    (with cost set to 1, net gain to sender) and alternative values
    (with cost set to 4, major loss to sender):

```{r}
x.low <- setx(z.out1, cost = 1)
x.high <- setx(z.out1, cost = 4)
```

    Simulate fitted values and first differences:

```{r}
s.out1 <- sim(z.out1, x = x.low, x1 = x.high)
# summary(s.out1)
plot(s.out1)
```

2.  `Joint Estimation of a Model with Different Sets of Explanatory
    Variables`[sto.dep.probit]

    Using the sample data `sanction`, estimate the statistical model,
    with `import` a function of `coop` in the
    first equation and `export` a function of
    `cost` and `target` in the second equation:
    
    
```{r}
fml2 <- list(mu1 = import ~ coop, mu2 = export ~ cost + target)
z.out2 <- zelig(fml2, model = "bprobit", data = sanction)
summary(z.out2)
```

    Set the explanatory variables to their means:
    
```{r}
x.out2 <- setx(z.out2)
```

    Simulate draws from the posterior distribution:
    
```{r}
s.out2 <- sim(z.out2, x = x.out2)
summary(s.out2)
# plot(s.out2)
```

3.  Joint Estimation of a Parametrically and Stochastically Dependent
    Model

    Using the sample data `sanction`. The bivariate model is
    parametrically dependent if $Y_1$ and $Y_2$ share some or all
    explanatory variables, `*and*` the effects of the shared
    explanatory variables are jointly estimated. For example,
    
    
```{r}
fml3 <- list(mu1 = import ~ tag(coop, "coop") +
               tag(cost, "cost") + + tag(target, "target"),
             mu2 = export ~ tag(coop, "coop") + tag(cost, "cost") + tag(target, "target"))
```

```{r}
z.out3 <- zelig(fml3, model = "bprobit", data = sanction)
summary(z.out3)
```

    Note that this model only returns one parameter estimate for each of
    `coop`, `cost`, and `target`.
    Contrast this to Example [basic.bp] which returns two parameter
    estimates for each of the explanatory variables. Set values for the
    explanatory variables:
    
```{r}
x.out3 <- setx(z.out3, cost = 1:4)
```

    Draw simulated expected values:
    
```{r}
s.out3 <- sim(z.out3, x = x.out3) RRR\> summary(s.out3)
```

## Model

For each observation, define two binary dependent variables, $Y_1$ and
$Y_2$, each of which take the value of either 0 or 1 (in the following,
we suppress the observation index $i$). We model the joint outcome
$(Y_1$, $Y_2)$ using two marginal probabilities for each dependent
variable, and the correlation parameter, which describes how the two
dependent variables are related.

-   The *stochastic component* is described by two latent (unobserved)
    continuous variables which follow the bivariate Normal distribution:

    $$\begin{aligned}
      \left ( \begin{array}{c} 
          Y_1^* \\
          Y_2^* 
        \end{array}
      \right ) &\sim &  
      N_2 \left \{ \left ( 
          \begin{array}{c}
            \mu_1 \\ \mu_2
          \end{array} \right ), \left( \begin{array}{cc}
                     1 & \rho \\
                     \rho & 1 
                     \end{array} \right) \right\},\end{aligned}$$

    where $\mu_j$ is a mean for $Y_j^*$ and $\rho$ is a scalar
    correlation parameter. The following observation mechanism links the
    observed dependent variables, $Y_j$, with these latent variables

    $$\begin{aligned}
    Y_j & = & \left \{ \begin{array}{cc}
                       1 & {\rm if} \; Y_j^* \ge 0, \\
                       0 & {\rm otherwise.}
                       \end{array} 
                       \right.\end{aligned}$$

-   The *systemic components* for each observation are

    $$\begin{aligned}
        \mu_j & = & x_{j} \beta_j \quad {\rm for} \quad j=1,2, \\
        \rho & = & \frac{\exp(x_3 \beta_3) - 1}{\exp(x_3 \beta_3) + 1}.\end{aligned}$$

## Quantities of Interest

For $n$ simulations, expected values form an $n \times 4$ matrix.

-   The expected values (`qi$ev`) for the binomial probit
    model are the predicted joint probabilities. Simulations of
    $\beta_1$, $\beta_2$, and $\beta_3$ (drawn form their sampling
    distributions) are substituted into the systematic components, to
    find simulations of the predicted joint probabilities
    $\pi_{rs}=\Pr(Y_1=r, Y_2=s)$:

    $$\begin{aligned}
    \pi_{11} &= \Pr(Y_1^* \geq 0 , Y_2^* \geq 0) &= \int_0^{\infty}
    \int_0^{\infty} \phi_2 (\mu_1, \mu_2, \rho) \, dY_2^*\, dY_1^* \\
    \pi_{10} &= \Pr(Y_1^* \geq 0 , Y_2^* < 0)  &= \int_0^{\infty}
    \int_{-\infty}^{0} \phi_2 (\mu_1, \mu_2, \rho) \, dY_2^*\, dY_1^*\\
    \pi_{01} &= \Pr(Y_1^* < 0 , Y_2^* \geq 0)  &= \int_{-\infty}^{0}
    \int_0^{\infty} \phi_2 (\mu_1, \mu_2, \rho) \, dY_2^*\, dY_1^*\\
    \pi_{11} &= \Pr(Y_1^* < 0 , Y_2^* < 0)  &= \int_{-\infty}^{0}
    \int_{-\infty}^{0} \phi_2 (\mu_1, \mu_2, \rho) \, dY_2^*\, dY_1^*\\\end{aligned}$$

    where $r$ and $s$ may take a value of either 0 or 1, $\phi_2$ is the
    bivariate Normal density.

-   The predicted values (`qi$pr`) are draws from the
    multinomial distribution given the expected joint probabilities.

-   The first difference (`qi$fd`) in each of the predicted
    joint probabilities are given by
    $$\textrm{FD}_{rs} = \Pr(Y_1=r, Y_2=s \mid x_1)-\Pr(Y_1=r, Y_2=s
      \mid x).$$

-   The risk ratio (`qi$rr`) for each of the predicted joint
    probabilities are given by

    $$\textrm{RR}_{rs} = \frac{\Pr(Y_1=r, Y_2=s \mid x_1)}{\Pr(Y_1=r, Y_2=s \mid x)}.$$

-   In conditional prediction models, the average expected treatment
    effect (`att.ev`) for the treatment group is

    $$\frac{1}{\sum_{i=1}^n t_i}\sum_{i:t_i=1}^n \left\{ Y_{ij}(t_i=1) -
          E[Y_{ij}(t_i=0)] \right\} \textrm{ for } j = 1,2,$$

    where $t_i$ is a binary explanatory variable defining the treatment
    ($t_i=1$) and control ($t_i=0$) groups. Variation in the simulations
    are due to uncertainty in simulating $E[Y_{ij}(t_i=0)]$, the
    counterfactual expected value of $Y_{ij}$ for observations in the
    treatment group, under the assumption that everything stays the same
    except that the treatment indicator is switched to $t_i=0$.

-   In conditional prediction models, the average predicted treatment
    effect (`att.pr`) for the treatment group is

    $$\frac{1}{\sum_{i=1}^n t_i}\sum_{i:t_i=1}^n \left\{ Y_{ij}(t_i=1) -
          \widehat{Y_{ij}(t_i=0)}\right\} \textrm{ for } j = 1,2,$$

    where $t_i$ is a binary explanatory variable defining the treatment
    ($t_i=1$) and control ($t_i=0$) groups. Variation in the simulations
    are due to uncertainty in simulating $\widehat{Y_{ij}(t_i=0)}$, the
    counterfactual predicted value of $Y_{ij}$ for observations in the
    treatment group, under the assumption that everything stays the same
    except that the treatment indicator is switched to $t_i=0$.

## Output Values

The output of each Zelig command contains useful information which you
may view. For example, if you run
`z.out <- zelig(y ~ x, model = bprobit, data)`, then you may examine the
available information in `z.out` by using `names(z.out)`, see the
`coefficients` by using `z.out$coefficients`, and
obtain a default summary of information through `summary(z.out)`. Other
elements available through the `$` operator are listed
below.

-   From the `zelig()` output object `z.out`, you
    may extract:

    -   `coefficients`: the named vector of coefficients.

    -   `fitted.values`: an $n \times 4$ matrix of the
        in-sample fitted values.

    -   `predictors`: an $n \times 3$ matrix of the linear
        predictors $x_j \beta_j$.

    -   `residuals`: an $n \times 3$ matrix of the residuals.

    -   `df.residual`: the residual degrees of freedom.

    -   `df.total`: the total degrees of freedom.

    -   `rss`: the residual sum of squares.

    -   `y`: an $n \times 2$ matrix of the dependent
        variables.

    -   `zelig.data`: the input data frame if `save.data
        = TRUE`.

-   From `summary(z.out)`, you may extract:

    -   `coef3`: a table of the coefficients with their
        associated standard errors and $t$-statistics.

    -   `cov.unscaled`: the variance-covariance matrix.

    -   `pearson.resid`: an $n \times 3$ matrix of the
        Pearson residuals.

-   From the `sim()` output object `s.out`, you
    may extract quantities of interest arranged as arrays indexed by
    simulation $\times$ quantity $\times$ `x`-observation
    (for more than one `x`-observation; otherwise the
    quantities are matrices). Available quantities are:

    -   `qi$ev`: the simulated expected values (joint
        predicted probabilities) for the specified values of
        `x`.

    -   `qi$pr`: the simulated predicted outcomes drawn from
        a distribution defined by the joint predicted probabilities.

    -   `qi$fd`: the simulated first difference in the
        predicted probabilities for the values specified in
        `x` and `x1`.

    -   `qi$rr`: the simulated risk ratio in the predicted
        probabilities for given `x` and `x1`.

    -   `qi$att.ev`: the simulated average expected
        treatment effect for the treated from conditional prediction
        models.

    -   `qi$att.pr`: the simulated average predicted
        treatment effect for the treated from conditional prediction
        models.

How to Cite {#how-to-cite .unnumbered}
-----------

See also {#see-also .unnumbered}
--------

The bivariate probit function is part of the VGAM package by Thomas Yee
[@YeeHas03]. In addition, advanced users may wish to refer to
`help(vglm)` in the VGAM library. Additional documentation is available
at .Sample data are from @Martin92
